Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-09-08 jso.qmd
practicalities section

<!--  
```{=html}
\begin{assumption}\label{assum:my_assumption}
This is an assumption.
\end{assumption}
Assumption \ref{assum:my_assumption} is convenient.

\begin{claim}\label{claim:my_claim}
This is a claim.
\end{claim}
Claim \ref{claim:my_claim} is suspicious.

\begin{condition}\label{cond:my_condition}
This is a condition.
\end{condition}
Condition \ref{cond:my_condition} is necessary.

\begin{conjecture}\label{conj:my_conjecture}
This is a conjecture.
\end{conjecture}
Conjecture \ref{conj:my_conjecture} is intriguing.

\begin{cor}\label{cor:my_corollary}
This is a corollary.
\end{cor}
Corollary \ref{cor:my_corollary} is trivial.

\begin{defn}\label{def:my_definition}
This is a definition.
\end{defn}
Definition \ref{def:my_definition} is clear.

\begin{example}\label{ex:my_example}
This is an example.
\end{example}
Example \ref{ex:my_example} is enlightening.

\begin{lem}\label{lem:my_lemma}
This is a lemma.
\end{lem}
Lemma \ref{lem:my_lemma} is helpful.

\begin{notation}\label{not:my_notation}
This is a notation.
\end{notation}
Notation \ref{not:my_notation} is cumbersome.

\begin{problem}\label{prob:my_problem}
This is a problem.
\end{problem}
Problem \ref{prob:my_problem} is difficult.

\begin{proof}\label{proof:my_proof}
This is a proof.
\end{proof}
Proof \ref{proof:my_proof} is compelling.

\begin{prop}\label{prop:my_prop}
This is a proposition.
\end{prop}
Proposition \ref{prop:my_prop} is useful.

\begin{rem}\label{rem:my_remark}
This is a remark.
\end{rem}
Remark \ref{rem:my_remark} is interesting.

\begin{thm}\label{thm:my_theorem}
This is a theorem.
\end{thm}
Theorem \ref{thm:my_theorem} is important.
```
-->

@@

@@

(not done yet)

The JSO can be used by specifying the argument `search_f = search_jellyfish` in the `guided_tour()` function in the `tourr` package. In the example below, the `lda_pp` index is used to cluster the penguins data set. When binding animation `animate_xy()` to an object (`res<-`), it will save the visited bases of each jellyfish as a data object. The bases from each jellyfish can be extracted adn supplied as a planned tour to visualise its path. Figure @fig-jso-penguins shows the tour path of one particular jellyfish in the penguins data set. The jellyfish successfully finds the structure of the data set, separating the three species of penguins.

```{r}
#| echo: true
#| eval: false
# run the projection pursuit with JSO 
res <- animate_xy(penguins_df, 
                  guided_tour(lda_pp(cl = penguins_cl$species),
                              search_f = search_jellyfish, 
                              n_jellies = 20, max.tries = 50))
# extract the bases from the first jellyfish
bases <- res |> filter(loop == 1) |> pull(basis) |> check_dup(0.1)
# visualise its path with a planned tour
animate_xy(data = penguins_df, tour_path = planned_tour(bases), 
           col = penguins_cl$species)
```

```{r}
#| label: fig-jso-penguins
#| fig-cap: "Visualising the tour path of one particular jellyfish in the penguins data set. The jellyfish is optimised by the JSO with 20 jellyfishes and a maximum of 50 tries. The tour path is planned off-line and then followed in the guided tour. The jellyfish successfully finds the structure of the data set, separating the three species of penguins."
frames <- c("001", "010", "019", "028", "036", "048")
ani <- paste0(here::here("figures/"), "tour/", "penguins", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl)
```

```{r}
#| eval: false
#| echo: false
jso_optima <- save_history(flea[,1:6], 
                           guided_tour(holes()), 
                           search_f = search_jellyfish)
```

>>>>

@@

To use JSO for PPGT in the `tourr` package, specify `search_f = search_jellyfish` in the guided tour. Since multiple tour paths are explored by the jellyfish, the animation function `animate_*()` will not render the tour path for JSO. To visualise path visited by each individual jellyfish, assign the animation to an object (`res <- animate_xy(...)`). This will save the bases visited by the optimiser, along with relevant metadata, as a tibble data object (see @RJ-2021-105 for more details on the data object). The bases visited for individual jellyfish can then be extracted and viewed as a planned tour. The following code uses JSO, with 20 jellyfish and a maximum of 50 tries, on the penguins data using the `lda_pp` index.

```{r}
#| echo: true
#| eval: false
# clean up the penguins data
penguins_cl <- palmerpenguins::penguins |> 
  filter(!is.na(bill_length_mm)) |>
  rename(bl = bill_length_mm, bd = bill_depth_mm,
         fl = flipper_length_mm, bm = body_mass_g)
penguins_df <- penguins_cl |> select(bl:bm) |> rescale()
# run the projection pursuit with JSO 
res <- animate_xy(penguins_df, 
                  guided_tour(lda_pp(cl = penguins_cl$species),
                              search_f = search_jellyfish, 
                              n_jellies = 20, max.tries = 50))
```

The bases visited by jellyfish No. 1 is extracted with `filter(loop = 1)` (the variable `loop` in the data object is the jellyfish ID) and the `check_dup()` function removes the bases in the iteration without improvements. The bases are then supplied as a `planned_tour()` in `animate_xy()` to visualise the tour path of jellyfish No.1.

```{r}
#| echo: true
#| eval: false
# extract the bases from the first jellyfish
bases <- res |> filter(loop == 1) |> pull(basis) |> check_dup(0.1)
# visualise its path with a planned tour
animate_xy(data = penguins_df, tour_path = planned_tour(bases), 
           col = penguins_cl$species)
```

@fig-jso-penguins shows six frames from the tour jellyfish No.1 animation. The Gentoo species (red) was mixed with Chinstrap (yellow) initially and gradually separated out as flipper length (`fl`) gaining more weights.

```{r}
#| label: fig-jso-penguins
#| fig-cap: "Visualising the tour path of a single jellyfish (No.1) in the penguins data set. The projection pursuit is optimised by the JSO with 20 jellyfishes and a maximum of 50 tries. The bases visited by jellyfish No.1 is extracted and constructed as a planned tour. The jellyfish successfully finds the structure of the data set, separating the three species of penguins."
frames <- c("001", "010", "019", "028", "036", "048")
ani <- paste0(here::here("figures/"), "tour/", "penguins", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl)
```

@@

### 2024-09-08 jso.tex
practicalities section

\begin{tcolorbox}[enhanced jigsaw, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, arc=.35mm, colback=white, bottomtitle=1mm, opacityback=0, breakable, bottomrule=.15mm, rightrule=.15mm, left=2mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, coltitle=black, leftrule=.75mm]

@@

@@

(not done yet)

The JSO can be used by specifying the argument
\texttt{search\_f\ =\ search\_jellyfish} in the \texttt{guided\_tour()}
function in the \texttt{tourr} package. In the example below, the
\texttt{lda\_pp} index is used to cluster the penguins data set. When
binding animation \texttt{animate\_xy()} to an object
(\texttt{res\textless{}-}), it will save the visited bases of each
jellyfish as a data object. The bases from each jellyfish can be
extracted adn supplied as a planned tour to visualise its path. Figure
Figure~\ref{fig-jso-penguins} shows the tour path of one particular
jellyfish in the penguins data set. The jellyfish successfully finds the
structure of the data set, separating the three species of penguins.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# run the projection pursuit with JSO }
\NormalTok{res }\OtherTok{\textless{}{-}} \FunctionTok{animate\_xy}\NormalTok{(penguins\_df, }
                  \FunctionTok{guided\_tour}\NormalTok{(}\FunctionTok{lda\_pp}\NormalTok{(}\AttributeTok{cl =}\NormalTok{ penguins\_cl}\SpecialCharTok{$}\NormalTok{species),}
                              \AttributeTok{search\_f =}\NormalTok{ search\_jellyfish, }
                              \AttributeTok{n\_jellies =} \DecValTok{20}\NormalTok{, }\AttributeTok{max.tries =} \DecValTok{50}\NormalTok{))}
\CommentTok{\# extract the bases from the first jellyfish}
\NormalTok{bases }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{|\textgreater{}} \FunctionTok{filter}\NormalTok{(loop }\SpecialCharTok{==} \DecValTok{1}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{pull}\NormalTok{(basis) }\SpecialCharTok{|\textgreater{}} \FunctionTok{check\_dup}\NormalTok{(}\FloatTok{0.1}\NormalTok{)}
\CommentTok{\# visualise its path with a planned tour}
\FunctionTok{animate\_xy}\NormalTok{(}\AttributeTok{data =}\NormalTok{ penguins\_df, }\AttributeTok{tour\_path =} \FunctionTok{planned\_tour}\NormalTok{(bases), }
           \AttributeTok{col =}\NormalTok{ penguins\_cl}\SpecialCharTok{$}\NormalTok{species)}
\end{Highlighting}
\end{Shaded}

\caption{\label{fig-jso-penguins}Visualising the tour path of one
particular jellyfish in the penguins data set. The jellyfish is
optimised by the JSO with 20 jellyfishes and a maximum of 50 tries. The
tour path is planned off-line and then followed in the guided tour. The
jellyfish successfully finds the structure of the data set, separating
the three species of penguins.}

>>>>

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, colback=white, breakable, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, opacityback=0, titlerule=0mm, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, rightrule=.15mm, toprule=.15mm, left=2mm]

@@

To use JSO for PPGT in the \texttt{tourr} package, specify
\texttt{search\_f\ =\ search\_jellyfish} in the guided tour. Since
multiple tour paths are explored by the jellyfish, the animation
function \texttt{animate\_*()} will not render the tour path for JSO. To
visualise path visited by each individual jellyfish, assign the
animation to an object (\texttt{res\ \textless{}-\ animate\_xy(...)}).
This will save the bases visited by the optimiser, along with relevant
metadata, as a tibble data object (see Zhang et al. (2021) for more
details on the data object). The bases visited for individual jellyfish
can then be extracted and viewed as a planned tour. The following code
uses JSO, with 20 jellyfish and a maximum of 50 tries, on the penguins
data using the \texttt{lda\_pp} index.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# clean up the penguins data}
\NormalTok{penguins\_cl }\OtherTok{\textless{}{-}}\NormalTok{ palmerpenguins}\SpecialCharTok{::}\NormalTok{penguins }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{filter}\NormalTok{(}\SpecialCharTok{!}\FunctionTok{is.na}\NormalTok{(bill\_length\_mm)) }\SpecialCharTok{|\textgreater{}}
  \FunctionTok{rename}\NormalTok{(}\AttributeTok{bl =}\NormalTok{ bill\_length\_mm, }\AttributeTok{bd =}\NormalTok{ bill\_depth\_mm,}
         \AttributeTok{fl =}\NormalTok{ flipper\_length\_mm, }\AttributeTok{bm =}\NormalTok{ body\_mass\_g)}
\NormalTok{penguins\_df }\OtherTok{\textless{}{-}}\NormalTok{ penguins\_cl }\SpecialCharTok{|\textgreater{}} \FunctionTok{select}\NormalTok{(bl}\SpecialCharTok{:}\NormalTok{bm) }\SpecialCharTok{|\textgreater{}} \FunctionTok{rescale}\NormalTok{()}
\CommentTok{\# run the projection pursuit with JSO }
\NormalTok{res }\OtherTok{\textless{}{-}} \FunctionTok{animate\_xy}\NormalTok{(penguins\_df, }
                  \FunctionTok{guided\_tour}\NormalTok{(}\FunctionTok{lda\_pp}\NormalTok{(}\AttributeTok{cl =}\NormalTok{ penguins\_cl}\SpecialCharTok{$}\NormalTok{species),}
                              \AttributeTok{search\_f =}\NormalTok{ search\_jellyfish, }
                              \AttributeTok{n\_jellies =} \DecValTok{20}\NormalTok{, }\AttributeTok{max.tries =} \DecValTok{50}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

The bases visited by jellyfish No.~1 is extracted with
\texttt{filter(loop\ =\ 1)} (the variable \texttt{loop} in the data
object is the jellyfish ID) and the \texttt{check\_dup()} function
removes the bases in the iteration without improvements. The bases are
then supplied as a \texttt{planned\_tour()} in \texttt{animate\_xy()} to
visualise the tour path of jellyfish No.1.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# extract the bases from the first jellyfish}
\NormalTok{bases }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{|\textgreater{}} \FunctionTok{filter}\NormalTok{(loop }\SpecialCharTok{==} \DecValTok{1}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{pull}\NormalTok{(basis) }\SpecialCharTok{|\textgreater{}} \FunctionTok{check\_dup}\NormalTok{(}\FloatTok{0.1}\NormalTok{)}
\CommentTok{\# visualise its path with a planned tour}
\FunctionTok{animate\_xy}\NormalTok{(}\AttributeTok{data =}\NormalTok{ penguins\_df, }\AttributeTok{tour\_path =} \FunctionTok{planned\_tour}\NormalTok{(bases), }
           \AttributeTok{col =}\NormalTok{ penguins\_cl}\SpecialCharTok{$}\NormalTok{species)}
\end{Highlighting}
\end{Shaded}

Figure~\ref{fig-jso-penguins} shows six frames from the tour jellyfish
No.1 animation. The Gentoo species (red) was mixed with Chinstrap
(yellow) initially and gradually separated out as flipper length
(\texttt{fl}) gaining more weights.

\caption{\label{fig-jso-penguins}Visualising the tour path of a single
jellyfish (No.1) in the penguins data set. The projection pursuit is
optimised by the JSO with 20 jellyfishes and a maximum of 50 tries. The
bases visited by jellyfish No.1 is extracted and constructed as a
planned tour. The jellyfish successfully finds the structure of the data
set, separating the three species of penguins.}

@@
